Social inequality in mortality among adults and elderly in northern Sweden 1851--2013\footnote{Paper presented at the session 'Health and Mortality in Arctic Populations'. The Ninth International Congress on Arctic Social Sciences, Umeå, June 8--12, 2017}

### 2017-05-31 icass.Rnw
Revision of diverse sections. Adding references

\date{May 30, 2017}

@@

\begin{abstract}
A long-term perspective on the development of social
inequalities in mortality in the adult and elderly population is taken in
this paper.
The area under investigation consists of the Skellefteå and Umeå regions in
the north of Sweden, and the time period is 1851--2013. The outcome is all
cause mortality and cause specific mortality for the categories
cardiovascular diseases and cancers. The analysis of inequality is based on
Hisclass.

@@

One of the central aspects of survival is social class and access to economic and other resources. Even in present welfare societies, social position is a strong determinant when it comes to health and mortality and the impact even seems to be increasing \citep{kuetal04, frlu07}. It has been suggested that “… social conditions have been, are and will continue to be irreducible determinants of health outcomes and therefore deserves their appellation as ‘fundamental causes’ of disease and death” \citep{liph95}. For a long time, the general view has been that socio-economic health inequalities were high in historical societies, probably higher than in modern societies. This is a reasonable assumption since these societies were in most cases characterised by very large differences between groups and classes. Knowing that access to resources provides advantages in all ways, the health advantage of higher classes ought to be obvious. \citet{ant67} and \citet{smi83} suggest that social inequality in health has passed through different historical phases. According to them, social health differences were comparatively small during the pre-transitional phase when space was a strong determinant for the spread of disease. Differences then increased during the transitional phase when mortality declined and wealthy groups used their resources to gain better health. Finally, health differences decreased again in modern low-mortality settings when instead health-related behaviour became important, resulting in marginal differences. \citet{omran82} state, in the third proposition in his theory of the epidemiological transition, that even if the class differential in mortality was maintained during the transition, the decline set in earlier and was faster among privileged groups.

@@

Another question of the development of social inequality in mortality, concerns its relation to economic inequality. \citet{wipi09} argue that income inequality has an independent effect on mortality, separate from the direct effect of actual different access to economic resources. They argue that unequal societies perform less well when it comes to health than equal ones. Wilkinson and Pickett find a positive association between inequality and health (as well as other social conditions) at country level in the present-day economically developed world. This has initiated a vital scholarly debate and the topic has been extensively studied \citep{suka04, wado00}. How the association looked like in previous periods is unknown. This is of particular interest since the levels of economic inequality were very different from those of the recent decades. Even if it not a necessary implication, we would still assume that poorer groups were those most disadvantaged of large inequality in itself. The conditions during the 19th century are uncertain, but at least during the late 19th and early 20th century both income and wealth distribution was strongly skewed in Sweden. From a couple of decades into the 20th century, income and wealth inequality continuously diminished to reach a low point at around 1980. Thereafter, economic inequality has increased substantially until the present, although far from being as high as a century ago \citep{rowa08, rowa09}. The Swedish development resembles that of other European and North American countries in its basic features \citep{rowa14}.

@@

In the present paper, we take a long-term perspective on the development of social inequalities in mortality in the adult and elderly population. The areas under investigation are the Skellefteå and Umeå regions in the north of Sweden, and the time period is 1851--2013 for Skellefteå and 1901--2013 for Umeå. The main questions are:

\begin{enumerate}
\item Has inequality in mortality between social classes increased among the adult and elderly
  population, primarily benefiting the highest social positions and how is it connected to the development of economic inequality in society?

%%\item Is the mortality transition initiated by a certain social class?

\item Are there any gender differences in the effect of social position?

\item Is social position equally important among the retired population as for those in working age?
\item Is there a difference between classes in mortality from cardiovascular diseases and cancers respectively?
\item Do class inequalities in survival correspond to those depending on level of education and income categories for the period 1990 on-wards?
\end{enumerate}

@@

The Skellefteå and Umeå regions (Figure~\ref{fig:mapp}) are part of the county of Västerbotten in the north of Sweden along the coast of the Gulf of Bothnia. This was a remote part of the country, where communication with the rest of Sweden was difficult until the late 19th century. The economy was dominated by agriculture, making it vulnerable to harvest failures, and several severe famines affected the regions during the 1800s. One living long in memory was the famine of 1867–-1868 \citep{segb14}. During the long winters, sea communication was hindered due to the Gulf of Bothnia being frozen. The first ships during the year came to this area in May or even as late as June in some cases \citep{fah56}. Towards the end of the 19th century, the Swedish railway system reached this part of Sweden, thereby facilitating contacts with the rest of Sweden, improving the economy and mitigating the effects of harvest failures. The inland was particularly remote and isolated. During the 19th century, our regions became increasingly integrated in the same epidemiological pattern as the rest of Sweden. Diseases reached to places that earlier were spared.

@@

Mortality was fairly low in comparison with other parts of the country and the fertility transition was late. The Skellefteå population size as defined in our data sets was 16\,473 on January 1, 1850, 41\,352 on January 1, 1900, 61\,938 on January 1, 1950 and 76\,723 at the end of the century.

The Umeå region in this study consists of Umeå urban and rural parish until 1950, and from 1960 on-wards of Umeå municipality when another three parishes are included. This region has a somewhat different character from that of Skellefteå. Umeå town had a small population but was substantially larger than Skellefteå town during almost the whole studied period. It was the administrative centre in the county of Västerbotten. Schools and military regiments were placed here and the economy was more diversified compared to Skellefteå. Agriculture dominated the rural part, but there were also some foundries as well as industries for example in forestry and small-scale production. The population size as defined in our data sets was 18\,970 on January 1, 1900, 32\,900 on January 1, 1950 and 103\,970 when the century ended.

@@

The recent extension in time of data at the Demographic Data Base, Umeå University (DDB) by including the period between 1900 and 1950s makes it possible to investigate the long-term development of social inequality in mortality, in the present case from 1851 to 2013. The databases provide us with micro-data on a sufficiently large population and with substantial social diversity for the Skellefteå and Umeå regions \citep{wee16}. The data for the present study comes from two large population databases at the DDB, Umeå University. The early period until 1950 is covered by
the database Poplink, the digitisation of historical parish registers for
the Skellefteå and Umeå region \citep{wee16}. Poplink is based on linked parish records, allowing
us to reconstruct life biographies on people as long as they remained in
the region. The records are linked within but not between the regions. Data
from Poplink are accessed for the period 1851--1950.

The other large data-set is extracted from the Linnaeus database
\citep{malm10}. Linnaeus is based on different linked population registers from
1960 to 2013 and is used within the ageing program at CEDAR, Umeå
University. The study period from 1960 to 2000 are constructed from censuses every fifth
year 1960--1990, with additional information on deaths from National Board of Health and Welfare. For the period 2002--2013 we use the information from the yearly population registers (LISA) together with death information from National Board of Health and Welfare.

@@

In the data-set analysed here, all individuals 40 years and older ever being resident in either
of the regions are included. The data file contains the variables on sex,
birth date, death date, first and last date of observation and type of
entrance/exit. Total number of persons are 84\,599 leading to 35\,001 deaths in Poplink and 197\,864 persons leading to 66\,297 deaths in the Linnaeus database. See Figure~\ref{fig:lexis2}.

@@

For the last period, 2002--2013, we use the yearly population registers (also known as LISA data from Statistics Sweden), together with information on deaths from National Board of Health and Welfare. This data allows good control of the presence periods in the studied regions.

@@

The availability to information on occupations varies over time and between sources. Occupations in Poplink are coded into HISCO \citep{lema02} and the original DDB coding system. The censuses 1960--1990 include occupational codes based on the system used in Sweden at that time, NYK (Nordisk Yrkesklassificering). The yearly population registers (LISA) do not report occupations until 2001, thus there is a window of 11 years from 1991 to 2001 where we lack information. In the censuses, we have solved this by extending the periods for which we consider the observed social class to be valid. Thus, the social class in the census 1990 is considered to be kept until 2001. This strategy is also used for other cases of missing social class. Furthermore, we consider the social position after retirement (in Sweden usually at age 65) to be the same as the last during working age.

@@

For the Poplink data, we have used a script re-coding HISCO into HISCLASS,
provided to us by Jonas Helgertz, Lund University. The procedure for
occupations in the Linnaeus database is a little more complicated. The
occupations in the censuses (NYK) are first transferred to the system
later in use, SSYK (the Swedish version of ISCO88). Thereafter the
SSYK, both those transferred from the censuses and those given in the
population registers 2001-2013, can be re-coded into HISCLASS.

Figures~\ref{fig:hiscskumm} and \ref{fig:hiscskumw} show the distribution (per cent) of exposure
time according to class for men and women respectively. The category of workers is
fairly stable over time. The highest class has
increased substantially, although from a very low level in the 19th
century. The increase of academics in connection to the establishment of
Umeå University explains much of this increase. What is not illustrated in Figure~\ref{fig:hiscskumm} is the transfer within the middle group from a majority of farmers during the 19th century to lower white collar from the middle of the 20th century. Furthermore we observe an increase during the 1960s and partly during later decades as well in those where social class has not been possible to define. When first observed in the earliest included censuses, occupation is often missing. For men the majority are however in the age group 65 and older, i.e. those that were retired. Until 1950, there were however very few with missing occupation. For men in working-active age, the proportion missing does not cause any problems for our analyses.

@@

<<hiscskummyoung, fig.cap = "Exposure by hisclass, men 40-65 (from the top: elite, middle, workers, missing) and time period.", fig.height = 4.5>>=
xx <- with(vb0Tab[vb0Tab$sex == "male" & vb0Tab$age %in% c("40", "45", "50", "55", "60"), ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
yy <- round(prop.table(xx, 1) * 100, 1)
barplot(t(yy)[4:1, ], col = c("red", "blue", "green", "black"), las = 2, ylab = "Per cent")
@

<<hiscskumwyoung, fig.cap = "Exposure by hisclass, women 40-65 (from the top: elite, middle, workers, missing) and time period.", fig.height = 4.5>>=
xx <- with(vb0Tab[vb0Tab$sex == "female" & vb0Tab$age %in% c("40", "45", "50", "55", "60"), ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
yy <- round(prop.table(xx, 1) * 100, 1)
barplot(t(yy)[4:1, ], col = c("red", "blue", "green", "black"), las = 2, ylab = "Per cent")
@

The proportion with missing values among women is however a larger problem, see Figure~\ref{fig:hiscskumw}. Corresponding to what we found for men, the proportion without social position is very low until the middle of the 20th century. This is a reflection of how womens social position is defined for that period. Missing social class is however a real problem during the 1960s and 1970s in particular. Around 80 \% lack information for the 1960s and 50 \% for the 1970s. Since social class for this period is defined according to own occupation, the consequence of many women being outside the workforce, the proportion missing becomes very high. We have chosen not to report results for women for these two decades. For the later decades, the results are fairly complete and ought to be representative.

The corresponding numbers for the age group 40--65 are shown in Figures~\ref{fig:hiscskummyoung} and 
\ref{fig:hiscskumwyoung}. Slightly lower rates of missing occupation in comparison to Figures~\ref{fig:hiscskumm} and \ref{fig:hiscskumw} are noted.

@@

Marital status is treated as a time-dependent variable in the data file. The statuses used are unmarried, married and dissolved marriages (without distinguishing between widowed and divorces). The status unmarried is often not explicitly noted in the source, which makes the marital status unknown in the database Poplink for people for people moving into our parishes. If no explicit status is given in the sources and partner is missing, the status has been set to unmarried.

Regarding the distribution of marital status over time, see Figure~\ref{fig:civstskum}.The figure indicate an under-registration of those married and and over-registration of previously married during the 1960s.

@@

@@

We analyse both regions together but perform separate analyses for men and
women and for the age groups 40--64 and 65--89. The model in the analyses
include social class, sex and marital status. Each decade
during the 20th century gets separate analyses, but we then combine results
in order to show the development of the social patterning of mortality
during the studied period. We present the results for the period 2001--2013
separately. We suggest a heterogeneity index in order to investigate a
possible homogenisation of mortality between social classes.

The important explanatory variable, {\sc Hisclass}, is included in the
proportional hazards models as a \emph{stratification} variable. The main
reason for this choice is that the effect of {\sc Hisclass} on survival is
non-proportional, that is, it varies with age. With the reduced data set
and Poisson regression, this is simply achieved by introducing an
interaction term between {\tt age} and {\tt hisclass}. Then, the main
results are
presented \emph{graphically}. There is some  lack of information in the higher
ages around 1960--1980 (see Figures~\ref{fig:hiscskumm} and \ref{fig:hiscskumw}), because that kind of 
information is simply missing to a large extent.

@@

Notice that we have not analysed the 1950s due to the fact that we only have data for some parishes at that time, consequently having no corresponding population for that decade. Furthermore, we have excluded the results for women during the 1960s and 1970 of reasons discussed above. The changing economy also have consequences for the analysis, particularly when it comes to the agricultural sector. While farmers were common occupations during the first half of the 20th century, these occupations has constantly decreased in numbers thereafter. This means that the middle class has somewhat different character during the last periods in the figures.

@@

A different pattern appears among men, being the most surprising finding in our analyses. The groups we would expect to have the lowest mortality, if we assume that access to different resources and
having a high status determines survival, instead have by far the highest hazards of dying during the nineteenth and much of the 20th century. This is also in accordance with our previous studies
\citep{edli11, edbr12}. Workers on the other hand have comparatively low mortality until the 1960s, at the same level as the middle class. The relative positions change radically towards the end the 20th century. From the 1980s, there is a clear advantage when it comes to survival for higher
managers and professionals. The relative differences are however more compressed than for the period until 1950. The expected social gradient in mortality has developed. This is also what we usually find in analyses of social health inequalities in present-day Sweden as well as in most
other countries. In the regions we are studying, this is however a quite
recent phenomenon. Corresponding to what we find for women, those without known occupation have much higher mortality than the others from the 1960s onwards. The assumption made for women of unknowns social class is probably even more relevant for men. They probably represent groups in society that are more vulnerable when it comes to mortality.

@@

The general pattern and development of mortality (not shown here) do not differ that much from the younger age group, of course representing a much higher mortality level earlier in history. Mortality among women was, as expected, lower than among men. We find a clear decline in mortality, but the fasted decline did not set in definitely until sometimes after the Second World War (results not shown here), i.e. later than for the age group 40--64 years. This corresponds to the Swedish development where we find a turning point at this time with better survival among elderly, to a large extent caused by lower mortality in cardiovascular diseases.

Concerning the social pattern, the pattern is less clear compared to what we found for the younger age group (see Figure~\ref{fig:cumswold}). For women the results are mixed. Once again, too much missing data on social class leaves out the results for the 1960s and 1970s. We do not find the same clear advantage for élite women during period until 1950, but neither do we find any large disadvantage for working class elderly women at that time. In some periods, the social differences were small, and the élite even had substantially higher mortality during the early 20th century. Notice however that this group was rather small at the time. The social differences continued to be small during the last decades of the 20th century. Only during the two last periods (after the year 2000) is it possible to establish the expected social gradient in mortality.

For men we find indications of the same intriguing change among elderly as for
the adult male opulation 40--64 years. Mortality was in fact substantially
higher among élite men  but their disadvantage was primarily apparent during the 19th century. Differences are fairly small during the first half of the 20th century. Not until the second half of the 20th century is the survival much better for the highest social class while workers in the age of retirement had high mortality in comparison. ). This is quite unexpected, since other studies have shown clear socio-economic differences in Sweden during recent decades.

@@

In this section, the focus is on analyses of mortality for the two dominating causes of death among the adult and elderly population. We start with cardiovascular diseases. Many studies have shown that socio-economic conditions are strongly associated with mortality in this disease group. Belonging to working class, having less education and low income increases the mortality risks. We would therefore expect to find a clear social pattern in the present analysis.

First, a general observation is that the analysis of causes of death and their development are more difficult to interpret, partly because the number of cases are fewer but also that the quality and coverage of reported death causes differ. Thus, many deaths during the early 20th got quite unspecific causes reported in the death registers. There was still a substantial part of all deaths where cause was diagnosed as unknown or dying from old age, despite the fact that reporting causes of death had become mandatory. For the period 1901--1925 to 1926--1950, the proportion of deaths from cardiovascular diseases increased from 25 to 43 per cent for the age group 40 and above, thus indicating an increase of these diseases. However, at the same time unknown cause and death of old age decreased from 28 percent to 18 per cent. It is quite reasonable to suspect that many cardiovascular deaths were disguised in unspecific categories and the increase therefore is explained by improvements in how deaths were diagnosed.
<<causes41>>=
cvb <- vb0
cvb$event <- cvb$heart ## Can this be done better?
cvb <- cvb[!(cvb$period %in% c("1851-1875", "1876-1900")), ]
cvb$period <- factor(cvb$period)

@@

As before, models are fitted separately for sex and each ten-year time period, from  1901 to
2013 and . We stratify on HISCLASS
and use the covariates urban/rural and {\tt   civst}. The cumulative
hazards at the end of the time period (65 if 40--64, 90 if 65--89) are the
main target in our analysis: These numbers are used in the process of
graphically illustrating the development over time.

@@

Performing the same types of analyses as for all-cause mortality, we start by looking at cumulative hazards by {\sc Hisclass} at age 65, given survival to age 40. They are adjusted for {\tt marital status}. We present separate analyses for women and men.

@@

Figure~\ref{fig:41cumsw} shows mortality levels during different periods from 1901 on-wards. Women from the highest social class had consistently lower mortality from cardiovascular diseases compared to the others. We also observe increasing social differences from the first half of the century to the last decades. A new pattern has developed from 1980 onwards. Now working class women have substantially higher cardiovascular compared to others. We can however not exclude that this is an artefact of different diagnostics for deaths in different groups. Working class women may have had less precise diagnoses in the early 20th century.

>>>>

\date{May 31, 2017}

@@

\begin{abstract}
A long-term perspective on the development of social
inequalities in mortality in the adult and elderly population is taken in
this paper. The area under investigation consists of the Skellefteå and Umeå regions in
the north of Sweden, and the time period is 1851--2013. The outcome is all
cause mortality and cause specific mortality for the categories
cardiovascular diseases and cancers. The analysis of inequality is based on
three social classes, reorganized from Hisclass.

@@

One of the central aspects of survival is social class and access to economic and other resources. Even in present-day welfare societies, social position is a strong determinant when it comes to health and mortality and the impact even seems to be increasing \citep{kuetal04, frlu07}. It has been suggested that “… social conditions have been, are and will continue to be irreducible determinants of health outcomes and therefore deserves their appellation as ‘fundamental causes’ of disease and death” \citep{liph95}. For a long time, the general view has been that socio-economic health inequalities were high in historical societies, probably higher than in modern societies. This is a reasonable assumption since these societies were in most cases characterised by very large differences between groups and classes. Knowing that access to resources provides advantages in all ways, the health advantage of higher classes ought to be obvious. \citet{ant67} and \citet{smi83} suggest that social inequality in health has passed through different historical phases. According to them, social health differences were comparatively small during the pre-transitional phase when space was a strong determinant for the spread of disease. Differences then increased during the transitional phase when mortality declined and wealthy groups used their resources to gain better health. Finally, health differences decreased again in modern low-mortality settings when instead health-related behaviour became important, resulting in marginal differences. \citet{omran82} state, in the third proposition in his theory of the epidemiological transition, that even if the class differential in mortality was maintained during the transition, the decline set in earlier and was faster among privileged groups.

@@

Another question of the development of social inequality in mortality, concerns its relation to economic inequality. \citet{wipi09} argue that income inequality has an independent effect on mortality, separate from the direct effect of actual different access to economic resources. They argue that unequal societies perform less well when it comes to health than equal ones. Wilkinson and Pickett find a positive association between inequality and health (as well as other social conditions) at country level in the present-day economically developed world. This has initiated a vital scholarly debate and the topic has been extensively studied \citep{suka04, wado00}. How the association looked like in previous periods is unknown. This is of particular interest since the levels of economic inequality were very different from those of the recent decades. Even if it not a necessary implication, we would still assume that poorer groups were those most disadvantaged of large inequality in itself. The situation during the 19th century is uncertain, but at least during the late 19th and early 20th century both income and wealth distribution was strongly skewed in Sweden. From a couple of decades into the 20th century, income and wealth inequality continuously diminished to reach a low point at around 1980. Thereafter, economic inequality has increased substantially until the present, although far from being as high as a century ago \citep{rowa08, rowa09}. The Swedish development resembles that of other European and North American countries in its basic features \citep{rowa14}.

@@

In the present paper, we take a long-term perspective on the development of social inequalities in mortality in the adult and elderly population. The areas under investigation are the Skellefteå and Umeå regions in the north of Sweden, and the time period is 1851--2013 for Skellefteå and 1901--2013 for Umeå. The main question we investigate is if inequality in mortality between social classes increased among the adult and elderly population, primarily benefiting the highest social positions and how this is connected to the development of economic inequality in society. This issue is then analysed according to gender, age-group (working age vs retired) and cause of death (cardiovascular and cancer). Furthermore, we compare our results witn analyses of alternative inequality indicators, education and income for the period 1986--2013.

% ? :
% 
% \begin{enumerate}
% \item Has inequality in mortality between social classes increased among the adult and elderly
%   population, primarily benefiting the highest social positions and how is it connected to the development of economic inequality in society?
% 
% %%\item Is the mortality transition initiated by a certain social class?
% 
% \item Are there any gender differences in the effect of social position?
% 
% \item Is social position equally important among the retired population as for those in working age?
% \item Is there a difference between classes in mortality from cardiovascular diseases and cancers respectively?
% \item Do class inequalities in survival correspond to those depending on level of education and income categories for the period 1990 on-wards?
% \end{enumerate}
% 
%

@@

The Skellefteå and Umeå regions (Figure~\ref{fig:mapp}) are part of the county of Västerbotten in the north of Sweden along the coast of Gulf of Bothnia. This was a remote part of the country, where communication with the rest of Sweden was difficult until the late 19th century. The economy was dominated by agriculture, making it vulnerable to harvest failures, and several severe famines affected the regions during the 1800s. One living long in memory was the famine of 1867–-1868 \citep{segb14}. During the long winters, sea communication was hindered due to the Gulf of Bothnia being frozen. The first ships during the year came to this area in May or even as late as June in some cases \citep{fah56}. Towards the end of the 19th century, the Swedish railway system reached this part of Sweden, thereby facilitating contacts with the rest of Sweden, improving the economy and mitigating the effects of harvest failures. The inland was particularly remote and isolated. During the 19th century, our regions became increasingly integrated in the same epidemiological pattern as the rest of Sweden. Diseases reached to places that earlier were spared.

@@

Mortality was fairly low in comparison with other parts of the country and the fertility transition was late \citep{coa86}. The Skellefteå population size as defined in our data sets was 16\,473 on January 1, 1850, 41\,352 on January 1, 1900, 61\,938 on January 1, 1950 and 76\,723 at the end of the century.

The Umeå region in this study consists of Umeå urban and rural parish until 1950, and from 1960 onwards of Umeå municipality when another three parishes are included. This region has a somewhat different character from that of Skellefteå. Umeå town had a small population but was substantially larger than Skellefteå town during almost the whole studied period. It was the administrative centre in the county of Västerbotten. Schools and military regiments were placed here and the economy was more diversified compared to Skellefteå. Agriculture dominated the rural part, but there were also some foundries as well as industries for example in forestry and small-scale production. The population size as defined in our data sets was 18\,970 on January 1, 1900, 32\,900 on January 1, 1950 and 103\,970 when the century ended.

@@

The recent extension in time of data at the Demographic Data Base (DDB), Umeå University (www.cedar.umu.se) by including the period between 1900 and 1950s makes it possible to investigate the long-term development of social inequality in mortality, in the present case from 1851 to 2013. The databases provide us with micro-data on a sufficiently large population and with substantial social diversity for the Skellefteå and Umeå regions \citep{wee16}. The data for the present study comes from two large population databases at the DDB. The early period until 1950 is covered by
the database Poplink, the digitisation of historical parish registers for
the Skellefteå and Umeå region \citep{wee16}. Poplink is based on linked parish records, allowing
us to reconstruct life biographies on people as long as they remained in
the region. The records are linked within but not between the regions. Data
from Poplink are accessed for the period 1851--1950.

The other large data-set is extracted from the Linnaeus database
\citep{malm10}. Linnaeus is based on different linked national population registers from
1960 to 2013 and is used within the ageing program at CEDAR, Umeå
University. The study period from 1960 to 2000 are constructed from censuses every fifth
year 1960--1990, with additional information on deaths from National Board of Health and Welfare. For the period 2002--2013 we use the information from the yearly population registers from Statistics Sweden (LISA) together with death information from National Board of Health and Welfare.

@@

In the data-set analysed here, all individuals 40 years and older ever being resident in either
of the regions are included. The data file contains the variables on social class (HISCLASS) sex, urban/rural residence, birth date, death date, first and last date of observation and type of
entrance/exit. Total number of persons are 84\,599 leading to 35\,001 deaths in Poplink and 197\,864 persons leading to 66\,297 deaths in the Linnaeus database. See Figure~\ref{fig:lexis2}.

@@

For the last period, 2002--2013, we use the yearly population registers (LISA data), together with information on deaths from National Board of Health and Welfare. This data allows good control of the presence periods in the studied regions.

@@

The availability to information on occupations varies over time and between sources. Occupations in Poplink are coded into HISCO \citep{lema02} and the original DDB coding system. The censuses 1960--1990 include occupational codes based on the system used in Sweden at that time, NYK (Nordisk Yrkesklassificering). The yearly population registers (LISA) do not report occupations until 2001, thus there is a window of 11 years from 1991 to 2001 where we lack information. In the censuses, we have solved this by extending the periods for which we consider the observed social class to be valid. Thus, the social class in the census 1990 is considered to be kept until 2001. This strategy is also used for other cases of missing social class. Furthermore, we consider the last social position before retirement to be kept until death or last observation for those above 65. For the Linnaeus data, we use a script recoding HISCO into HISCLASS, provided to us by Jonas Helgertz, Lund University.

@@

% The procedure for
% occupations in the Linnaeus database is a little more complicated. The
% occupations in the censuses (NYK) are first transferred to the system
% later in use, SSYK (the Swedish version of ISCO88). Thereafter the
% SSYK, both those transferred from the censuses and those given in the
% population registers 2001-2013, can be re-coded into HISCLASS.

Figures~\ref{fig:hiscskumm} and \ref{fig:hiscskumw} show the distribution (per cent) of exposure
time according to class for men and women respectively. The category of workers is
fairly stable over time. The highest class has increased substantially, although from a very low level in the 19th century. The increase of academics in connection to the establishment of
Umeå University explains much of this increase. What is hidden in Figure~\ref{fig:hiscskumm} is the transfer within the middle group from a majority of farmers during the 19th century to lower white collar from the middle of the 20th century. Furthermore we observe an increase during the 1960s and partly during later decades as well in those with undefined. When first observed in the earliest included censuses, occupation is often missing. For men the proportion is somewhat higher in the age group 65 and above, i.e. those that were retired, but still low enough to be used in our analyses. Until 1950, there were however very few with missing occupation. For men in working-active age, the proportion missing does not cause any problems for our analyses.

@@

% <<hiscskummyoung, fig.cap = "Exposure by hisclass, men 40-65 (from the top: elite, middle, workers, missing) and time period.", fig.height = 4.5>>=
% xx <- with(vb0Tab[vb0Tab$sex == "male" & vb0Tab$age %in% c("40", "45", "50", "55", "60"), ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
% yy <- round(prop.table(xx, 1) * 100, 1)
% barplot(t(yy)[4:1, ], col = c("red", "blue", "green", "black"), las = 2, ylab = "Per cent")
% @
% 
% <<hiscskumwyoung, fig.cap = "Exposure by hisclass, women 40-65 (from the top: elite, middle, workers, missing) and time period.", fig.height = 4.5>>=
% xx <- with(vb0Tab[vb0Tab$sex == "female" & vb0Tab$age %in% c("40", "45", "50", "55", "60"), ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
% yy <- round(prop.table(xx, 1) * 100, 1)
% barplot(t(yy)[4:1, ], col = c("red", "blue", "green", "black"), las = 2, ylab = "Per cent")
% @

The proportion with missing values among women is however a larger problem, see Figure~\ref{fig:hiscskumw}. Corresponding to what we found for men, the proportion without social position is very low until the middle of the 20th century. This is a reflection of how womens social position is defined for that period. Missing social class is however a real problem during the 1960s and 1970s in particular. Around 80 \% lack information for the 1960s and 50 \% for the 1970s. Since social class for this period is defined according to own occupation, the consequence of many women being outside the workforce, the proportion missing becomes very high. We have chosen not to report results for women for these two decades. For the later decades, the results are fairly complete and should be representative.

% The corresponding numbers for the age group 40--65 are shown in Figures~\ref{fig:hiscskummyoung} and 
% \ref{fig:hiscskumwyoung}. Slightly lower rates of missing occupation in comparison to Figures~\ref{fig:hiscskumm} and \ref{fig:hiscskumw} are noted.

@@

Marital status is treated as a time-dependent variable in the data file. The statuses used are unmarried, married and dissolved marriages (without distinguishing between widowed and divorces). If no explicit status is given in the sources and partner is missing, the status has been set to unmarried.

Regarding the distribution of marital status over time, see Figure~\ref{fig:civstskum}. The figure indicates an under-registration of those married and and over-registration of previously married during the 1960s.

@@

\subsection{Urban vs rural residence}
In 19th and early 20th century Sweden (as in other countries), mortality was much higher in urban environments. The unsanitary and overcrowded living conditions as well as the risks for importing infections made the health risks higher in towns and cities. In the present paper, we control for this aspect by distinguishing between urban and rural residence, based on in which parish people lived.

@@

We analyse both regions together but perform separate analyses for men and
women and for the age groups 40--64 and 65--89. The model in the analyses
include social class, sex and marital status. Each decade
during the 20th century gets separate analyses, but we then combine results
in order to show the development of the social patterning of mortality
during the studied period.

% We present the results for the period 2001--2013
% separately. We suggest a heterogeneity index in order to investigate a
% possible homogenisation of mortality between social classes.

The important explanatory variable, {\sc Hisclass}, is included in the
proportional hazards models as a \emph{stratification} variable. The main
reason for this choice is that the effect of {\sc Hisclass} on survival is
non-proportional, that is, it varies with age. With the reduced data set
and Poisson regression, this is simply achieved by introducing an
interaction term between {\tt age} and {\tt hisclass}. Then, the main
results are presented \emph{graphically}.

% There is some  lack of information in the higher
% ages around 1960--1980 (see Figures~\ref{fig:hiscskumm} and \ref{fig:hiscskumw}), because that kind of 
% information is simply missing to a large extent.

@@

Notice that we have not analysed the 1950s due to the fact that we only have data for some parishes at that time, consequently having no corresponding population for that decade. Furthermore, we have excluded the results for women during the 1960s and 1970 of reasons discussed above. The changing economy also have consequences for the analysis, particularly when it comes to the agricultural sector. While farmers were common occupations during the first half of the 20th century, these occupations has constantly decreased in numbers thereafter. This means that the middle class has somewhat different character during the last periods.

@@

A different pattern appears among men, being the most surprising finding in our analyses. The groups we would expect to have the lowest mortality, if we assume that access to different resources and
having a high status determines survival, instead have by far the highest hazards of dying during the nineteenth and much of the 20th century. This is also in accordance with our previous studies
\citep{edli11, edbr12}. Workers on the other hand have comparatively low mortality until the 1960s, at the same level as the middle class. The relative positions change radically towards the end the 20th century. From the 1980s, there is a clear advantage when it comes to survival for higher
managers and professionals. The relative differences are however more compressed than for the period until 1950. The expected social gradient in mortality has developed. This is also what we usually find in analyses of social health inequalities in present-day Sweden as well as in most
other countries. In the regions we are studying, this is however a quite
recent phenomenon. Corresponding to what we find for women, those without known occupation have much higher mortality than the others from the 1960s onwards. The assumption made for women of unknowns social class is probably even more relevant for men. They probably represent groups in society that are particularly vulnerable.

@@

The general pattern and development of mortality (not shown here) do not differ that much from the younger age group, of course representing a much higher mortality level earlier in history. Mortality among women was, as expected, lower than among men. We find a clear decline in mortality, but the fastest decline did not set in definitely until sometimes after the Second World War (results not shown here), i.e. later than for the age group 40--64 years. This corresponds to the Swedish development where we find a turning point at this time with better survival among elderly, to a large extent caused by lower mortality in cardiovascular diseases.

Concerning the social pattern, the pattern is less clear compared to what we found for the younger age group (see Figure~\ref{fig:cumswold}). For women the results are mixed. Once again, too much missing data on social class leaves out the results for the 1960s and 1970s. We do not find the same clear advantage for élite women during period until 1950, but neither do we find any large disadvantage for working class elderly women at that time. In some periods, the social differences were small, and the élite even had substantially higher mortality during the early 20th century. The social differences continued to be small during the last decades of the 20th century. Only during the two last periods (after the year 2000) is it possible to establish the expected social gradient in mortality.

For men we find indications of the same intriguing change among elderly as for
the adult male population 40--64 years. Mortality was in fact substantially
higher among élite men  but their disadvantage was primarily apparent during the 19th century. Differences are fairly small during the first half of the 20th century. Not until the second half of the 20th century is the survival much better for the highest social class while workers in the age of retirement had high mortality in comparison. ). This is quite unexpected, since other studies have shown clear socio-economic differences in Sweden during recent decades.

@@

In this section, the focus is on analyses of mortality from the two dominant causes of death among the adult and elderly population. We start with cardiovascular diseases. Many studies have shown that socio-economic conditions are strongly associated with mortality in this disease group. Belonging to working class, having less education and low income increases the mortality risks. We would therefore expect to find a clear social pattern in the present analysis.

First, a general observation is that the analysis of causes of death and its development is difficult to interpret, partly because the number of cases are fewer but also that the quality and coverage of reported death causes differ. Thus, many deaths during the early 20th got quite unspecific causes reported in the death registers. There was still a substantial part of all deaths where cause was diagnosed as unknown or dying from old age, despite the fact that reporting causes of death had become mandatory. For the period 1901--1925 to 1926--1950, the proportion of deaths from cardiovascular diseases increased from 25 to 43 per cent for the age group 40 and above, thus indicating an increase of these diseases. However, at the same time unknown cause and death of old age decreased from 28 percent to 18 per cent. It is quite reasonable to suspect that many cardiovascular deaths were disguised in unspecific categories and that the increase partly is explained by improvements in how deaths were diagnosed.
<<causes41>>=
cvb <- vb0
cvb$event <- cvb$heart ## Can this be done better?
cvb <- cvb[!(cvb$period %in% c("1851-1875", "1876-1900")), ]
cvb$period <- factor(cvb$period)

@@

As before, models are fitted separately for sex and each ten-year time period, from  1901 to
2013. We stratify on HISCLASS and use the covariates urban/rural and {\tt   civst}. The cumulative
hazards at the end of the time period (65 if 40--64, 90 if 65--89) are the
main target in our analysis: These numbers are used in the process of graphically illustrating the development over time.

@@

Performing the same types of analyses as for all-cause mortality, we start by looking at cumulative hazards by {\sc Hisclass} at age 65, given survival to age 40. They are adjusted for {\tt marital status} and urban/rural residence. We present separate analyses for women and men.

@@

Figure~\ref{fig:41cumsw} shows mortality levels during different periods from 1901 on-wards. Women from the highest social class had consistently lower mortality from cardiovascular diseases compared to the others. We also observe increasing social differences from the first half of the century to the last decades. A new pattern has developed from 1980 onwards. Now working class women have substantially higher cardiovascular mortality compared to others. We can however not exclude that this is an artefact of different diagnostics for deaths in different groups. Working class women may have had less precise diagnoses in the early 20th century.

For men, there is a different pattern, a pattern however that to a large extent is consistent with what we observed concerning all-cause mortality. The elite group had much higher cardiovascular diseases up until the middle of the 20th century, but instead have the lowest mortality from 1980 onwards. Working class men had a fairly low cardiovascular mortality, but in the same way as among women, the levels were fairly similar to the one of the middle category consisting of lower manager/farmers. And in the same way as for women at this age, working class men have the highest mortality from this cause of death during the last decades. A social hierarchy in cardiovascular mortality seems to have developed both for men and women.

% A word of caution is however needed when it comes to these results. We can not exclude that the smaller differences in cardiovascular diseases for women in the early periods are an artefact of how deaths were diagnosed. Working class men and women may have had less precise diagnoses in the early 20th century.
